---
title: "Assessing the Impact of *L. rhamnosus* Intervention on *Rumicoccus gnavus* Abundance in the Gut"
author: "Jess Cunliffe"
date: "May, 2024"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
# GLOBAL KNITR OPTIONS ----
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)
```

```{r loading packages}
library(tidyverse)
library(janitor)
library(rstatix)
library(performance)
library(here)
library(tidyr)
library(dplyr)
library(kableExtra)
library(GGally)
library(emmeans)
library(corrplot)
library(lmtest)
library(ggridges)
library(rmarkdown)
library(knitr)
library(patchwork)
```

```{r data reading, echo=FALSE, include=FALSE}
source("scripts/probiotic_3_analysis.R")
```

## Introduction

The human gut microbiome refers to the diverse community of
microorganisms that inhabit the gastrointestinal tract (GI). The
relationship between the gut microbiome and its host is multifaceted and
dynamic, with certain interactions playing a crucial and beneficial role
in digestion, nutrient metabolism, immune regulation, and protection
against pathogens. However, imbalance or dysregulation of the gut
microbiome has association with a wide range of health conditions,
including gastrointestinal, immune-related, and metabolic disorders
**(Kuziel and Rakoff-Nahoum, 2022)**. *Ruminococcus gnavus*, a
gram-positive anaerobic bacterium, is a common constituent of the
indigenous human gut microbiota. *R. gnavus* has been linked both
positively and negatively to an array of GI conditions. It is proven
that *R. gnavus* produces bacteriocins including *Ruminococcins*,
antimicrobial peptides that combat harmful bacteria, potentially
contributing to gut microbial balance. Conversely, dysbiosis of the gut
microbiota and increased abundance of *R. gnavus* is frequently
correlated with gut disorders, including inflammatory bowel disease
(IBD), colorectal cancer (CRC), and irritable bowel syndrome (IBS),
highlighting the pathogenic potential of this bacterium **(Emmanuelle
Crost et al., 2023).**

Another well-established gut bacterium is *Lactobacillus rhamnosus*
Gorbach-Goldin (GG), which is recognised for its probiotic properties
**(Papizadeh et al., 2016)**. The term ‘probiotics’ incorporates all
non-pathogenic living microorganisms that when administered in adequate
amounts, confer health benefits to the host **(Córdoba and Hotel,
2001)**. *L. rhamnosus* GG achieves this host benefit with robust
pili-appendages that endure challenging conditions enabling adherence to
intestinal mucus glycoproteins and colonisation of the gut. *L.
rhamnosus* GG is associated with reducing the risk of gestational
diabetes mellitus, improving immune reactions following vaccines, and
managing diarrhoea associated with cancer and antibiotic treatments.
Additionally, *L. rhamnosus* GG inhibits the growth and adherence of
various pathogenic bacteria and may help restore healthy gut microbiota
in conditions like IBS **(Steele, 2022).** Although both bacteria have
been extensively studied, there remains a considerable gap in research
regarding their relationship and potential interaction. It is this
study’s aim to examine whether the intervention treatment with *L.
rhamnosus* GG can result in a decrease in the abundance of *R. gnavus*
in human stool samples.

## Analysis

To study the potential interaction between *L. rhamnosus* GG and *R.
gnavus*, a sample of 22 male and female subjects produced two stool
samples at two-time points: a before and after sample. The subjects were
treated with either *L. rhamnosus* GG probiotic supplementation or a
placebo (denoted in the study by “group”). From the stool samples,
high-through profiling data was obtained to analyse the composition and
abundance of the bacterial genera in the intestines. The data was
analysed using RStudio Pro 2023.12.1 and packages such as tidyverse
**(Wickham et al, 2019)** ggplot2 **(Wickham, 2016)**, performance
**(Ludecke et al, 2021)**, kableExtra **(Zhu, 2024)** and “rmarkdown”
**(Allaire et al, 2020)** were used throughout. These packages allowed
for efficient and professional cleaning, statistical analysis, graphical
visualisation, and summary tables of the data to be produced.

### Data Cleaning  
  
  
Firstly, the data was manipulated by removing the columns: “subject” and
“sample”, which were insignificant for exploring a relationship. The
data underwent additional processing using pivot_wider to restructure it
into a more informative format, displaying abundance values for both
before and after observations for each subject. Then, a "difference"
column was generated using the mutate function, representing the
calculated values obtained by subtracting before abundance from after
abundance. The headers of the remaining columns were then put in
snake_case naming style for consistency.

### Data Analysis 
  
  
Utilising histograms and boxplots as visual aids, potential outliers
were explored, revealing a before abundance value of *R. gnavus* of 913.
This outlier was subsequently confirmed through a car::qqPlot analysis.
However, to comprehensively grasp the impact of this outlier on the
dataset, a model needed to be fitted.

### Data Modelling 

A least-squares linear model was used to investigate whether the
difference in *R. gnavus* abundance is affected by group treatment and
gender as categorical predictors. The model produced an insignificant
*F*-statistic and an adjusted *R*-squared value of -0.03 which indicates
that the model was a poor fit for the data. A Shapiro Wilk test
**(Shapiro and Wilk, 1965)** revealed that the residuals did not follow
a normal distribution (*W* = 0.88, *P* \< 0.05) and finally, a Cooks
Distance Plot **(Cook, 2011)** concluded the outlier was causing skew on
the data set. The outlier and its corresponding data points for the
subject were removed to address this issue, and a new least-squares
linear model was created which can be seen in **Table 1** (*F*(2,18)=
4.14, *P* \< 0.05, *R\^2* = 0.32). The Breusch-Pagan test (**Breusch and
Pagan, 1979)**, Shapiro-Wilk test and performance checks found
homoscedasticity and normal residuals in the data, confirming a good
fit. To explore gender as a potential interaction, an *F*-test was
conducted, comparing models with and without the interaction term. The
analysis revealed no significant interaction between the variables, yet
the model significantly improved with the inclusion of ‘gender’ so was
kept in for analysis. The final sample population was 21 participant and
quantitative information is summarised in **Table 2.**
  
  
```{r Table 1: Linear Model Summary , echo=FALSE}
model_3_summary %>% 
  kbl(digits = 2, caption = "Linear Model of R. gnavus Abundance Difference: Effects of Gender and Treatment Group") %>%
  kable_styling(bootstrap_options = "striped", full_width = T, position = "left")

```
  
  
```{r Table 2: Participant Summary , echo=FALSE}

probiotic_3_participant_summary <- probiotic_3 %>% 
  group_by(group, gender) %>% 
  summarise(quantity = n())
probiotic_3_participant_summary %>% 
  kbl(caption="Participant Information and Quantities", bold = TRUE) %>% 
  kable_styling(bootstrap_options = "striped", full_width = T, position = "left")

```

## Results

It was hypothesised that the abundance of *R.gnavus* in the gut is
reduced when intervened with *L. rhamnosus* GG supplementation. To test
this, a least squares linear model was used with treatment group and
gender as explanatory variables and the change in abundance between pre-
and post-treatment observations as the response variable. The results
provide intriguing insights into the potential of probiotic
supplementation and are summarised visually in both **Figure 1** and
**Figure 2.**

```{r Figure 1:Abundance Difference, echo=FALSE, fig.cap="Figure 1: Bar Chart containing raw data values of before and after abundance counts of R. gnavus. (Top Left) Both treatment groups result in an increase in abundance, LGG treatment resulting in an average increase of 37.43 ± 30.01 and the Placebo treatment resulting in an average increase of 34.93 ¬±24.49, showing very little effect of treatment group on abundance variance. (Top right) Shows a very small increase in R. gnavus abundance between before and after observation in Females but a large mean increase of 110 ¬± 27.60 in Males. (Bottom Graph) Combines all variables to show no effect of probiotic intervention in Females and an enhancing effect on R. gnavus abundance in Males.", fig.align = "center", fig.width = 10.7, fig.height= 9.5, out.width = "83%"}

((bar_1 + bar_2)/ bar_3)
```

```{r Figure 2, echo=FALSE, fig.cap="Figure 2: Changes in R. gnavus Abundance Pre- and Post-Treatment by Group and Gender. Top Graph: This graph depicts the mean changes in R. gnavus abundance between each sample by gender and treatment group. Females treated with L. rhamnosus GG show an average decrease of -4.20 ± 18.64, while those given a placebo show an average increase of 11.20 ± 27.92. The red dashed line indicates the average difference in abundance for each treatment group without gender differentiation, highlighting a lack of overall treatment effect but revealing gender disparities, as indicated by the pink (Female) and blue (Male) lines diverging from the average. These lines link the genders between each treatment group and the gradient between Placebo and LGG shows the impact the probiotic has on R. gnavus abundance within the gender. Males treated with LGG show an average increase of 141.50 ± 16.5, whereas those given a placebo show an increase of 94.25 ± 40.13. This suggests that the probiotic treatment has a more significant impact on R. gnavus abundance in males than in females.Bottom Graph: This graph illustrates the data distribution and interquartile ranges for each gender within the treatment groups, providing additional context on variability and spread.", fig.align = "center", fig.width = 8, fig.height= 10, out.width = "87%"}

(plot_1 / plot_2)

```

### *L. rhamnosus* GG Probiotic intervention impact on *R. gnavus* abundance   

As seen in **Table 3**, the result show that on average, the abundance
of *R. gnavus* increases by 34.93±24.49 in the Placebo group and
contradictory to the hypothesis, an increase of 37.43±30.01 when treated
with *L. rhamnosus* GG. The linear coefficient model in **Table 1**,
provides an estimate of -2.50 for the Placebo group which is found to be
an insignificant difference between treatment groups with a *P*-value of
0.94. Therefore, there is insufficient evidence to conclude that the
intervention of the *L. rhamnosus* GG has any effect on *R. gnavus*
abundance.


```{r Table 3: Summary of Group , echo=FALSE}

probiotic_3_summary_group <- probiotic_3 %>% 
  group_by(group) %>% 
  summarise(mean_difference_in_rcg_abund = mean(difference),
    sd = sd(difference),
    se = sd(difference) / sqrt(n()))
probiotic_3_summary_group %>% 
  kbl(caption="Summary of Mean Difference in R. gnavus Abundance by Treatment Group") %>% 
  kable_styling(bootstrap_options = "striped", full_width = T, position = "left")

```

### Gender impact on *R. gnavus* abundance 

Interestingly, **Figures 1 and 2** and **Table 4** reveal an unexpected
gender disparity between the two treatment groups. The linear model
confirms this, showing a significant increase of 103.93 in *R. gnavus*
abundance in males compared to females (p = 0.01). As seen in **Table
5**, *R. gnavus* abundance shows minimal change with either treatment
intervention or placebo in females, evidenced by a mean decrease of 4.20
± 18.64 in the *L. rhamnosus* GG group. This decrease is not significant
(*P* = 0.80), indicating that the probiotic does not significantly
affect *R. gnavus* abundance in females. Conversely, males show a
substantial increase in mean *R. gnavus* abundance in both treatment
groups, which contradicts the hypothesis. Specifically, there is a mean
increase of 141.50 ± 16.50 in the *L. rhamnosus* GG group and 94.25 ±
40.13 in the placebo group. While the model indicates a significant
gender effect on R. gnavus abundance, the Placebo variable is not
significant (*P* = 0.94), suggesting no significant difference in *R.
gnavus* abundance between the *L. rhamnosus* GG and placebo groups after
accounting for gender differences.


```{r Table 4: Summary of Gender , echo=FALSE}
 
probiotic_3_summary_gender <- probiotic_3 %>% 
  group_by(gender) %>% 
  summarise(mean_difference_in_rcg_abund = mean(difference),
            sd = sd(difference),
            se = sd(difference) / sqrt(n()))
probiotic_3_summary_gender %>% 
  kbl(caption="Summary of Mean Difference in R. gnavus Abundance by Gender") %>% 
  kable_styling(bootstrap_options = "striped", full_width = T, position = "left")

```

```{r Table 5: Summary of Group and Gender, echo=FALSE}

probiotic_3_summary <- probiotic_3 %>% 
  group_by(group,gender) %>% 
  summarise(mean_difference_in_rcg_abund = mean(difference),
            sd = sd(difference),
            se = sd(difference) / sqrt(n()))
probiotic_3_summary %>% 
  kbl(caption="Summary of Mean Difference in R. gnavus Abundance by Treatment Group and Gender") %>% 
  kable_styling(bootstrap_options = "striped", full_width = T, position = "left")

```


### Alternative Explanations 

The unexpected gender-specific responses in *R. gnavus* abundance could
be due to several factors. Baseline differences in gut microbiota,
hormonal influences, and variations in diet and lifestyle between males
and females may have contributed. It has been proved that women with
higher oestrogen levels correlated with increased diversity, higher
abundance of *Bacteroidetes* and fewer *Firmicutes*, while women with
raised testosterone levels exhibit less *Ruminococcus* bacteria. Men on
the other hand, with higher testosterone, are positively correlated with
*Ruminococcus* and microbial diversity **(d’Afflitto et al., 2022).**
Additionally, the probiotic strain's unique effects, immune system
modulation, and the complex microbial interactions within the gut
microbiota might have played a role the effectiveness of probiotic
supplementation. This gender disparity highlights the need for
gender-stratified analyses in microbiota research and the development of
precision probiotics tailored to individual microbiota profiles.
Advances in microbiome sequencing and bioinformatics can help design
probiotics targeting specific microbial imbalances in males and females.

### Limitations and Future Research 

While this study provides valuable insights into the gender-specific
effects of *L. rhamnosus* GG on *R. gnavus* abundance, it has
limitations such as a small sample size, short follow-up, and lack of
detailed metadata on diet, medication, and lifestyle. Future research
should address these by conducting larger, longitudinal studies with
detailed clinical phenotyping, multi-omics profiling, and species-level
analysis, which could elucidate the precise effects of *L. rhamnosus* GG
on *R. gnavus* and other gut microbes. Integrating microbiome data with
clinical, dietary, and lifestyle information can enhance understanding
of probiotic efficacy and personalised treatment responses.

## Conclusion

This study found no significant effect of *L. rhamnosus* GG on *R.
gnavus* abundance in the gut. However, a significant gender disparity
was observed: males exhibited a substantial increase in *R. gnavus*
abundance regardless of treatment, while females showed minimal change.
These findings suggest that gender-specific factors may influence gut
microbiota responses to probiotic interventions. Further research with
larger, more detailed studies is needed to understand these dynamics
better.

## Reference List

Allaire J, Xie Y, Dervieux C, McPherson J, Luraschi J, Ushey K, Atkins
A, Wickham H, Cheng J, Chang W, Iannone R (2024).\_rmarkdown: Dynamic
Documents for R\_. R package version 2.26’
<https://github.com/rstudio/rmarkdown>.

Breusch, T S & Pagan, A R, 1979. "A Simple Test for Heteroscedasticity
and Random Coefficient Variation," Econometrica, Econometric Society,
vol. 47(5), pages 1287-1294, September.

Cook, R.D. (2011). Cook’s Distance. In: Lovric, M. (eds) International
Encyclopedia of Statistical Science. Springer, Berlin, Heidelberg.
<https://doi.org/10.1007/978-3-642-04898-2_189>

Córdoba, A. and Hotel, P. (2001). ‘Health and Nutritional Properties of
Probiotics in Food including Powder Milk with Live Lactic Acid Bacteria
Report of a Joint FAO/WHO Expert Consultation on Evaluation of Health
and Nutritional Properties of Probiotics in Food Including Powder Milk
with Live Lactic Acid Bacteria Food and Agriculture Organization of the
United Nations World Health Organization Joint FAO/WHO Expert
Consultation on Evaluation of Health and Nutritional Properties of
Probiotics in Food Including Powder Milk with Live Lactic Acid
Bacteria’. [online] Available at:
<http://pc.ilele.hk/public/pdf/20190225/bd3689dfc2fd663bb36def1b672ce0a4.pdf>.

Crost, E., Coletto, E., Bell, A. and Juge, N. (2023). ‘Ruminococcus
gnavus: friend or foe for human health’. Fems Microbiology Reviews,
[online] 47(2). <doi:https://doi.org/10.1093/femsre/fuad014>.

d’Afflitto, M., Upadhyaya, A., Green, A. and Peiris, M. (2022).
‘Association Between Sex Hormone Levels and Gut Microbiota Composition
and Diversity—A Systematic Review’. Journal of Clinical
Gastroenterology, 56(5), pp.384–392.
<doi:https://doi.org/10.1097/mcg.0000000000001676>.

H. Wickham. ggplot2: Elegant Graphics for Data Analysis. Springer-Verlag
New York, 2016. Kuziel, G.A. and Rakoff-Nahoum, S. (2022). ‘The gut
microbiome’. Current Biology, 32(6), pp.R257–R264.
<doi:https://doi.org/10.1016/j.cub.2022.02.023>.

Lüdecke et al., (2021). performance: An R Package for Assessment,
Comparison and Testing of Statistical Models. Journal of Open Source
Software, 6(60), 3139. <https://doi.org/10.21105/joss.03139>

Papizadeh, M., Nahrevanian, H., Rohani, M., Hosseini, S.N. and
Shojaosadati, S.A. (2016). ‘Lactobacillus rhamnosus Gorbach-Goldin (GG):
A Top Well-Researched Probiotic Strain’. Journal of Medical
Bacteriology, [online] 5(5-6), pp.46–59. Available at:
<https://jmb.tums.ac.ir/index.php/jmb/article/view/318>.

SHAPIRO, S.S. & WILK, M.B., 1965. An analysis of variance test for
normality (complete samples). Biometrika, 52(3–4), pp.591–611. Available
at: <https://doi.org/10.1093/biomet/52.3-4.591>.

Steele, C. (2022). ‘Lactobacillus rhamnosus GG: A review of clinical use
and efficacy’. Nutr. Med. J., 1, 70-116.

Wickham H, Averick M, Bryan J, Chang W, McGowan LD, François R,
Grolemund G, Hayes A, Henry L, Hester J, Kuhn M, Pedersen TL, Miller E,
Bache SM, Müller, K, Ooms J, Robinson D, Seidel DP, Spinu V, Takahashi
K, Vaughan D, Wilke C, Woo, K, Yutani H (2019). “Welcome to the
tidyverse.” *Journal of Open Source Software*, *4*(43), 1686.
<doi:10.21105/joss.01686> <https://doi.org/10.21105/joss.01686>.

Zhu H (2024). *kableExtra: Construct Complex Table with 'kable' and Pipe
Syntax*. R package version 1.4.0,
<https://github.com/haozhu233/kableExtra>,
<http://haozhu233.github.io/kableExtra/>
