---
title: "Assessing the Impact of L. rhamnosus Intervention on Rumicoccus gnavus Abundance in the Gut"
author: "Jess Cunliffe"
date: "May, 2024"
output:
  pdf_document: 
    fig_caption: true 
    fig_height: 5  
    fig_width: 7 
---

```{r settings, include=FALSE}
# GLOBAL KNITR OPTIONS ----
knitr::opts_chunk$set(echo = FALSE,
                      fig.align = "center",
                      fig.width = 6,
                      fig.asp = 0.8,
                      out.width = "80%")
```

```{r setup, include=FALSE}
# PACKAGES ----
library(tidyverse)
library(janitor)
library(rstatix)
library(performance)
library(here)
library(tidyr)
library(dplyr)
library(kableExtra)
library(GGally)
library(emmeans)
library(corrplot)
library(lmtest)
library(ggridges)
library(rmarkdown)
library(knitr)
library(patchwork)
```

```{r read data, include=FALSE}
# READ DATA ----

probiotic <- read_csv(here("data", "probiotic.csv"))

source("scripts/probiotic_3_analysis.R")

```

## RMarkdown Notes

-   to initiate a new line, enter \*two spaces\*\* at the end of the previous line and then Enter/Return

-   Underscores (*text*) or single asterisk (*text*) to italicise

-   Double asterisks (**text**) for bold text

-   

    # First-level heading / Title

-   

    ## Second level heading

-   

    ### Third-level heading

-   echo = FALSE to not print the chunk’s R source code in the output document

-   warning = FALSE to not print warnings produced by the R code

-   message = FALSE to not print any messages produced by the R code

-   include = either TRUE/FALSE whether to include chunk outputs (e.g. plots) in the document

-   out.width = and out.height = - size of ouput e.g. out.width = "75%"

-   fig.align = "center" adjust how a figure is aligned across the page

-   fig.show='hold' if your chunk prints multiple figures and you want them printed next to each other (pair with out.width = c("33%", "67%")

## Introduction

**150-200 words**

\- Provide background information on the topic of the experiment.

\- State the purpose and objectives of the experiment.

\- Introduce relevant theories or concepts.

## Analysis Methods

**200-250 words**

\- Describe the materials used in the experiment, including equipment and chemicals.

\- Outline the methodology followed, including experimental procedures and data collection techniques.

\- Mention any controls or variables manipulated during the experiment.

```{r figure, echo = FALSE}




```

## Results and Discussion

**700 - 800 words**

\- Present the findings of the experiment in a clear and concise manner.

\- Include tables, graphs, or figures as necessary to support the results.

\- Describe any patterns or trends observed in the data.

\- Interpret the results of the experiment and discuss their significance.

\- Compare the findings to existing literature or theoretical expectations.

\- Analyze any limitations or sources of error in the experiment.

\- Propose possible explanations for unexpected results and suggest areas for further research.

```{r,echo =FALSE}
(plot_1+plot_2)+
  plot_layout(guides = "collect") 
```

```{r,echo =FALSE}
probiotic_3_participant_summary <- probiotic_3 %>% 
  group_by(group, gender) %>% 
  summarise(quantity = n())
probiotic_3_participant_summary %>% 
  kbl(caption="Table of Participant quantities within each Gender and Treatment Group") %>% 
  kable_styling(bootstrap_options = "striped", full_width = T, position = "left")
```

```{r,echo =FALSE}
probiotic_3_summary <- probiotic_3 %>% 
  group_by(group,gender) %>% 
  summarise(mean_difference_in_rcg_abund = mean(difference),
            sd = sd(difference),
            se = sd(difference) / sqrt(n()))
probiotic_3_summary %>% 
  kbl(caption="Summary statistics of gender and group compared with RCG abudance difference between before and after observations") %>% 
  kable_styling(bootstrap_options = "striped", full_width = T, position = "left")
```

```{r,echo =FALSE}
probiotic_3_summary_group <- probiotic_3 %>% 
  group_by(group) %>% 
  summarise(
    mean_difference_in_rcg_abund = mean(difference),
    sd = sd(difference),
    se = sd(difference) / sqrt(n()))
probiotic_3_summary_group %>% 
  kbl(caption="Summary statistics of treatment group compared with RCG abudance difference between before and after observations") %>% 
  kable_styling(bootstrap_options = "striped", full_width = T, position = "left")
```

```{r,echo =FALSE}
probiotic_3_summary_gender <- probiotic_3 %>% 
  group_by(gender) %>% 
  summarise(mean_difference_in_rcg_abund = mean(difference),
            sd = sd(difference),
            se = sd(difference) / sqrt(n()))
probiotic_3_summary_gender %>% 
  kbl(caption="Summary statistics of gender compared with RCG abudance difference between before and after observations") %>% 
  kable_styling(bootstrap_options = "striped", full_width = T, position = "left")
```

```{r, echo = FALSE}
model_3_summary <- summary(model_3)
model_3_summary <- tidy(model_3_summary)
kable(model_3_summary, format = "html", digits = 2) %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE)
```

## Conclusion

**100 words**

\- Summarize the key findings of the experiment.

\- Reiterate the main conclusions drawn from the results.

\- Discuss the broader implications of the findings and their relevance to the field of study.

## References
